MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators

### 2024-04-26 rebuttal.txt

- Extend literature in the intro (R1 Q1)
- More detailed background 
-- Add a new background subsection describing CPU impl (R1 Q2,3,7)
-- Elaborate more on MaBoSS output (perhaps a new subsection with plotted outputs) (R1 Q4,7 R2 Q1)
-- Revise the algorithm (R1 Q5,6)
- Revise GPU stuff (R1 Q8,10,11,13 R2 Q2,3.1,3.3,3.5)
- Revise MPI stuff (R2 Q3.2,3.4)
- in Results add a table with models we used ? (R1 Q9)
- explain the MPI outlier (remeasure?) (R1 Q12) 
- minor stuff (rest)

@@

First of all, we would like to thank the reviewers for their consideration of our paper and detailed remarks. We have addressed them in our paper revision. The individual answers are in the following. Minor remarks that were simply fixed in the paper and required no additional comments are omitted.

Reviewer #1:

Q: How is the parallelization for GPUs actually done? In the section MaBoSS.GPU, it is described, why there was no parallelization over the Boolean functions, but not about the rest. In the MPI section, it states, that each thread is assigned to compute a single trajectory at once. I would assume that this is also true for the GPU implementation, this should, however, be made clear.
A: You are right, we included this piece of information in the last paragraph of Simulation section.

@@

Q: There is no demonstration of how a user would actually deal with the code - especially since the GPU version is based on a hard-coded implementation of the rules. A demonstrative example would help to get an idea, how a user would actually get started, and how deep he needs to dig into the code - especially for a biological audience, accessability of such tools is an important point
A: Respectfully, it is not true that the GPU version is based on any hard-coded implementation. The compilation of the rules happens during the runtime without any user intervention. We included this note in the second paragraph of the Simulation section.

Reviewer #2:

Q: Figure 3: The x-axis states “Nodes count (log−scale)”, however, it appears to have a linear scale. The linear relationships claimed in the statement “We observed that the compilation time is linearly dependent on the number of nodes and formula lengths (measured in the number of occurring nodes).” is not clear from the provided figures, and should be clarified or, for example provide Figure 3 with a linear-linear or log-log scale.
A: Thanks for pointing out the wrong scale in the plot, it is a linear scale. Regarding the linear dependency, it can not be seen from Fig 3 since it shows compilation time percentages rather than the absolute values. Therefore, plotting a linear-linear scale would not show any linear dependency. Nevertheless, the stated linear dependency is indeed there and we observed it from the plots which we did not include in the paper. The general idea why it holds is that increasing the number of nodes by a value of x adds O(x) new lines to a source file which needs to be compiled. We included this explanation in the paper.

Q: I was able to successfully compile MaBoSS with MPI enabled, however I was not able to compile MaBoSS.GPU in a Linux environment.
A: Thanks, compiling with the newest CUDA version revealed a bug in the Thrust dependency. We committed a fix.

### 2024-04-26 Article.tex
some minor

In an idealized PRAM model with infinite parallelism, we can optimize the algorithm in the following ways:
\begin{itemize}
    \item Given $c$ processors, all trajectory simulations can be performed in parallel, reducing the time complexity to $\mathcal{O}(m \cdot n^2)$. (Note that this does not include the results aggregation. See \emph{Statistics aggregation} section for further description.)
    \item With $n$ processors, the computation of transition rates in the simulation can be done $\mathcal{O}(n)$ time, and the selection of the flipping bit can be done in $\mathcal{O}(\log{n})$ time using a parallel prefix sum, giving $\mathcal{O}(n)$ time for a single iteration.
\end{itemize}
Thus, using a perfect parallel machine with $c \cdot n$ processors, the computation time can be reduced to $\mathcal{O}(u \cdot n)$. Notably, the $\mathcal{O}(u)$ simulation steps that must be performed serially remain a major factor in the whole computation time.

@@

There are multiple ways to optimize the expression trees for GPUs: One may use a linked data structure that is more cache-friendly such as the van Emde Boas tree layout~\cite{van1975preserving}, or perhaps represent the Boolean formulas as a compact continuous array, or convert it to CNF or DNF bitmasks that can be easily evaluated by vector instructions. We decided to leave the exact representation choice on the compiler, by encoding the expressions as direct code and using the runtime compilation of GPU code~\cite{nvrtc}. In such an approach, the application reads the model files, writes the formulas as functions in CUDA C++, compiles them using the NVIDIA runtime compiler, and finally runs the simulation --- all without a user intervention.

@@

>>>>

In an idealized PRAM (parallel random access machine~\cite{fortune1978parallelism}) model with infinite parallelism, we can optimize the algorithm in the following ways:
\begin{itemize}
    \item Given $c$ processors, all trajectory simulations can be performed in parallel, reducing the time complexity to $\mathcal{O}(u \cdot n^2)$. (Note that this does not include the results aggregation. See \emph{Statistics aggregation} section for further description.)
    \item With $n$ processors, the computation of transition rates in the simulation can be done $\mathcal{O}(n)$ time, and the selection of the flipping bit can be done in $\mathcal{O}(\log{n})$ time using a parallel prefix sum, giving $\mathcal{O}(n)$ time for a single iteration.
\end{itemize}
Thus, using a perfect parallel machine with $c \cdot n$ processors, the computation time can be reduced to $\mathcal{O}(u \cdot n)$. Notably, the $\mathcal{O}(u)$ simulation steps that must be performed serially remain a major factor in the whole computation time.

@@

There are multiple ways to optimize the expression trees for GPUs: One may use a linked data structure that is more cache-friendly such as the van Emde Boas tree layout~\cite{van1975preserving}, or perhaps represent the Boolean formulas as a compact continuous array, or convert it to CNF or DNF (conjunctive or disjunctive normal form) bitmasks that can be easily evaluated by vector instructions. We decided to leave the exact representation choice on the compiler, by encoding the expressions as direct code and using the runtime compilation of GPU code~\cite{nvrtc}. In such an approach, the application reads the model files, writes the formulas as functions in CUDA C++, compiles them using the NVIDIA runtime compiler, and finally runs the simulation --- all without a user intervention.

@@

It is worth noting that the datacenter GPU performs worse than the laptop GPU. Both devices are bottlenecked by the runtime compilation of the Boolean formulas, however, NVIDIA A100 spends on average around $300$ms more on the compilation step. Subtracting the compilation time, A100 is faster for all models. We did not spend time finding the root cause of this discrepancy since the value is negligible and the following benchmarks show that the runtime compilation overhead quickly disappears with increasing model size.
